CRUSE to Safe Cycling in Ireland

### 2024-03-25 paper-v1.qmd

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography:
  # cruse.bib
  - bibliography.bib
  - cruse.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
  # html: default
  # docx: default
# # # # Uncomment for formatted version
  elsevier-pdf:
    linestretch: 2
    keep-tex: true
    journal:
      name: European Transport Research Review
      formatting: preprint
      model: 3p
      cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
author:
  - name: Robin Lovelace
    email: r.lovelace@leeds.ac.uk
    corresponding: true
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Joey Talbot
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Eugeni Vidal-Tortosa
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Hussein Mahfouz
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Elaine Brick
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Peter Wright
    affiliations:
      - name: AECOM
        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
  - name: "Gary O'Toole"
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Dan Brennan
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
  - name: Suzanne Meade
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
  - Transport Planning
  - Collaborative Planning
date: last-modified
---

@@

```{r}
#| include: false
library(tidyverse)
```

```{r}
#| include: false
#| eval: false
remotes::install_dev("tmap")
library(tmap)
# Save title page (1st page of paper.pdf):
system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
# Combine cv.pdf with visions.pdf:
system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
system("latexdiff paper-v1.tex paper.tex > tracked-changes.tex")
system("pdflatex tracked-changes.tex")
browseURL("tracked-changes.pdf")
```

@@

<!-- Mobility kills. -->
## Background

Transport systems dominated by heavy and powerful private cars are inefficient, dangerous, and unhealthy.
Cars are a major and growing source of greenhouse gas (GHG) emissions [@winkler2023], a leading cause of premature death and injury due to road traffic collisions [@globals2018], and a cause of disease due to physical inactivity and air, noise and microplastic pollution [@mattsson2023] [@welch2023] [@cavallaro2024]. <!-- Shouldn't these three sources share the brackets? --> Transport is responsible for 23% of GHG emissions, 70% of which are from road transport, with passenger cars accounting for nearly half of transport emissions (around 10% of global emissions) [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

Growing evidence of the negative impacts of car-dependent transport systems has led governments in many countries to set targets and take actions.
In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
The ASI framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.

Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient ebikes --- which are seeing rapid uptake --- outperform electric cars, which are too heavy and expensive, for the majority of trips.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
Climate change mitigation is a major motivation for cycle network plans [@scappini2022].

Another motivation for cycle network planning at the European level is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
Given that 'safety' in this context is usefully quantified as the number of people killed or seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
Better data to inform road safety policies and interventions is a motivation for better estimates of 'baseline' levels of physical activity at high levels of geographic resolution [@tait2023].

National governments are increasingly acting on the evidence.
In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
Doing so while simultaneously enabling rapid uptake of active modes will require key travel corridors to be identified 'cycle proofed'.
Cycling in Ireland represents only 3% of total modal share as of the 2016 Census, but accounts for 20% of serious injuries and 7% of all fatalities, driving the development of cycling policy and infrastructure plans in Ireland.
The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].

At the regional level within Ireland, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings of high support for and capability for cycling: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
Seven in ten trips in Ireland are by car [@national].
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

Further evidence of the importance of cycling in Ireland is provided by the National Strategic Objective (NSO) from the National Development Plan, which allocates €8.6 billion to sustainable transport infrastructure including public transport and active travel interventions.
Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.
It was in this context that Transport Infrastructure Ireland (TII) commissioned the research reported in this paper, which led to the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
Building on previous work, including the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.

Stakeholder consultation by TII emphasized the need for a strong, national, systematic but locally-specific evidence base on cycle networks in both urban and rural Ireland.
The evidence base had to scale nationally to support strategic alignment with national, regional and local policies, but also be useful for local network planning.

## Aim and content

This paper describes an approach to strategic cycle network planning tool development that is open, scalable, and evidence-based.
We label the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and present a case study of its application in Ireland, the results of which are publicly available at [cruse.bike](https://cruse.bike/).

In @sec-tools, we review existing current tools for estimating cycling potential.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss limitations and possible future improvements to the approach, and the implications of the results for Ireland before making concluding remarks in @sec-conclusions.

# Tools for estimating cycling potential {#sec-tools}

Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.
Some of the most relevant existing tools are summarised below and presented in @tbl-tools.

```{r}
#| include: false
tools = read_csv("tools.csv")
tools = tools |>
  transmute(
    `Approach` = paste0(Tool, ", ", Reference),
    `Coverage and input data` = paste0(Coverage, ", ", `Input data`),
    `Outputs and availability` = Output,
    License = Licence
  )
```

```{r}
#| label: tbl-tools
knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
  kableExtra::column_spec(1, width = "13em") |>
  kableExtra::column_spec(2, width = "15em") |>
  kableExtra::column_spec(3, width = "12em") |>
  kableExtra::column_spec(4, width = "5em")
```

Area-based tools were developed first.
Parkin and colleagues presented a regression model to estimate the percentage of cycling trips to work in England and Wales based on socioeconomic, transportation, and physical factors at the small area (wards) level in 2007 [@parkin_estimation_2007].
They found that areas with a higher percentage of females, non-whites, car ownership, lower socioeconomic classes, income, distance to work, population density, poor highway conditions, hills, rainfall, and fewer off-road bicycle routes tended to have a lower proportion of cycling to work.

Three years later, Transport for London created the Analysis of Cycling Potential (ACP) tool based on a mode choice model that assesses the likelihood of trips being cycled, with data from the TfL London Travel Demand Survey as its primary input.
The tool generates tables, graphs, and heatmaps showing the potential for cycling for different trips purposes in London to guide local cycling initiatives, such as where new hire 'docking stations' should go.
The tool does not use origin-destination data and cannot estimate cycling potential on specific routes [@transport_for_london_analysis_2010].

Other tools that generate areal data to prioritize investment include a study that generated a 'prioritization index' for Montreal, Canada [@larsen_build_2013], a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).

A more recent areal-based approach is the designed the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10.
They are then combined into an overall score, weighted by their impact on cycling according to the literature.
The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].

Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 @lovelace2017.
Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.

The approach presented in this paper seeks to overcome three limitations of previous tools to estimate cycling potential: 1) low resolution of data, with routes starting and ending in administrative zone centroids, 2) limited coverage of trip purposes beyond travel to work and school, and 3) a web interface that was not user-friendly or intuitive.

@@

```{=html}
<!-- 
The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionized strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].

@@

Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui. -->
```
<!-- Todo: add section on input datasets -->

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that data with high levels of geographic resolution are necessary to estimate cyclical potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
To tackle this issue we developed a new method for OD data disaggregation called 'jittering'. <!--This reference should be removed to anonymise the paper. They will guess that Lovelace is involved, but saying we developed and then citing is as if we did not remove authors names from the paper :) --> The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.

::: {#fig-dublin}
<!-- ![](images/paste-1.png) -->

![](images/paste-8.png) <!-- ![](images/paste-3.png) -->
![](images/paste-9.png)

Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland (botton).
Source: CRUSE Tool, publicly available at [cruse.bike](https://cruse.bike/).
:::

@@

However, travel to school and work only constitute around only 30% of all trips in Ireland. <!-- Reference needed -->
To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips. <!-- This is already implemented, right?--> 
An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes. <!--This should be Table 2--->

@@

While CycleStreets offers a free routing API, we commissioned a custom routing service to enable:

@@

We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the fastest routes minimize journey time, accounting for traffic lights and surface type.
The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
The 'Balanced' route type is a compromize between the two, minimising journey time while avoiding the busiest roads.
Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.

@@

These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
For the fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.

@@

Illustration of the quietest (left) and fastest (right) route networks for Naas (top) and Newbridge (bottom) in County Kildare.
:::

@@

The Climate Action Plan scenario models the transport emissions reductions targeted in the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which aims for a 51% cut in overall GHG emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.

To model this decrease in car km driven, cycling uptake increases in line with the Go Dutch scenario.
However, we only model shift from driving to cycling.
There is no shift from other modes of transport to cycling.

@@

::: {#fig-user-stories}
![](images/mermaid-figure-1.png)

Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.
:::

@@

## Comparison with other tools and contribution to the field

>>>>

---
title: "CRUSE to Safe Cycling in Ireland"
subtitle: An Open Source Tool to Support Active Travel
bibliography: bibliography.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
#   html: default
  # docx: default
# # # # Uncomment for formatted version
  elsevier-pdf:
    linestretch: 2
    keep-tex: true
    journal:
      name: European Transport Research Review
      formatting: preprint
      model: 3p
      cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
author:
  - name: Robin Lovelace
    email: r.lovelace@leeds.ac.uk
    corresponding: true
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Joey Talbot
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Eugeni Vidal-Tortosa
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Hussein Mahfouz
    affiliations:
      - name: University of Leeds
        address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
  - name: Elaine Brick
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Peter Wright
    affiliations:
      - name: AECOM
        address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
  - name: "Gary O'Tool"
    affiliations:
      - name: AECOM
        address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
  - name: Dan Brennan
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
  - name: Suzanne Meade
    affiliations:
      - name: Transport Infrastructure Ireland
        address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  Legislation mandates improved road safety and decarbonisation of predominantly road-based transport network in Ireland and other countries. Increasing cycling transport, making it safer and more attractive, is essential to this end. Cycling in Ireland represents only 3% of total modal share as of the latest Census (2016), but accounts for 20% of serious injuries and 7% of all fatalities. A planning tool based on high resolution datasets on baseline and potential cycling levels could greatly assist in achieving this goal.

  This paper describes the design, features and potential use of the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was commissioned to support cycle network planning, safety interventions, business cases and project evaluation. The resulting web application is available at <https://cruse.bike/>, enabling planners, engineers, and other stakeholders to make more evidence-based decisions.
  CRUSE goes beyond previous work in several important ways.
  It models networks at high levels of geographic resolution, simulates a wide range of trips including recreational, social, shopping and personal trips, in addition to trips to work and primary, secondary and tertiary education locations.
  The tool provides key 'denominator' information to enable reporting of collision rates per billion km cycled.

  Another key feature of CRUSE is its use of frequently-updated OpenStreetMap data for route calculation and "cycle friendliness" estimates of all links on the network.
  CRUSE provides three network types: the 'Fastest' network highlights routes for directness; 'Quietest' prioritises cycle friendliness, and 'Balanced' represents a balance between speed and comfort.
  Trips are generated based on origin and destination data from the 2016 Census in combination with modeled demand data to estimate current and potential future cycling levels at area, route, and network levels for each county in Ireland.

  For investment in cycling to deliver on the huge potential benefits, cycling must become safer.
  With growth in the ebike market, CRUSE will help inform inter-urban and rural networks to support the transfer of trips to sustainable modes for longer journeys.
  Provision of open access data in a publicly available web tool will address the challenges faced by counties across Ireland in meeting their sustainability commitments.
  The experience of developing a deploying CRUSE are relevant to countries with sustainable transport policies worldwide.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
date: last-modified
---

@@

```{r}
#| include: false
#| eval: false
library(tidyverse)
remotes::install_dev("tmap")
library(tmap)
# Save title page (1st page of paper.pdf):
system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
# Combine cv.pdf with visions.pdf:
system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
```

@@

<!-- Mobility kills. -->

High energy transport systems are a major contributor to climate change, a leading cause of premature death and injury due to road traffic collisions, and a cause of disease due to airborne and noise pollutants.
Conversely, evidence-based and effective transport *policies* have great potential to decarbonize economies, improve public health, and save lives.

Transport is responsible for 23% of global greenhouse gas (GHG) emissions, 70% of which are from road transport.
Nearly half of these emissions (around 10% of global emissions) are from passenger cars [@jaramillo2022].
The transport system encourages, enables and in some cases enforces unsustainable lifestyles.
Services that are only accessible by car lock-in car dependency [@gray2001; @shergold2012; @motte-baumvol2010].

Recognizing the growing evidence of such impacts of poorly designed and performing transport systems, governments in many countries have set targets and taken actions.
In the context of climate, road safety and physicial inactivity crises, policies to improve transport systems can be classified according to the 'Avoid-Shift-Improve' (ASI) framework [@jaramillo2022].
The ASI framework highlights the importance of demand reduction (*avoid*ing unnecessary trips), in addition to mode *shift* to sustainable modes and *improvement* of existing energy converters, in that order.

Building cycle networks represents a relatively 'quick win' within the context of decarbonization [@brand2020] and sustainable mobility [@burns2013].
Although cycling uptake appears on the surface to only relate to the 'shift' part of the ASI framework, closer consideration of the knock-on impacts of cycling uptake shows that it can also help avoid unnecessary trips [@nello-deakin2020].
Furthermore, highly efficient ebikes --- which are seeing rapid uptake in Ireland and other countries --- outperform electric cars, which are too heavy and expensive.
Over-reliance on electric cars could slow the transition away from car dependency and inadvertently enable "high travel lock-in" [@anable2019].
At the European level, the European Union has a target of reducing GHG emissions by 55% by 2030, compared to 1990 levels, and to achieve 'net-zero' by 2050 [@rosenow2022].
Climate change mitigation is a major motivation for cycle network plans [@scappini2022].

Another motivation for cycle network planning is the Road Infrastructure Safety Management (RISM) directive (2008/96/EC), which requires member states to implement a road safety management system (RSMS) for all public roads.
Specifically, "Member States shall ensure that the ranking of high accident concentration sections and the network safety ranking are carried out" [@directiv2008].
Given that 'safety' in this context is usefully quantified as the number of people killed and seriously injured (KSI) per distance traveled, the directive requires estimation of distance traveled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with motorized modes, this is a major challenge.
Better data to inform road safety policies and interventions is a motivation for the CRUSE Tool outlined in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.
A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].

At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car [@national] .
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as ebikes extend the range of this mode" [@greater].

Further evidence of the importance of cycling in Ireland *already* is provided by the National Strategic Objective (NSO) from the National Development Plan, with €8.6 billion allocated to sustainable transport infrastructure including public transport and active travel interventions.
Cycle infrastructure will be developed in synchrony with the BusConnects project, an entire redesign of the bus network in Dublin and Cork.

It was in this context that Transport Infrastructure Ireland (TII) commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
Based on the Propensity to Cycle Tool (PCT) for England and Wales, the CRUSE Tool was developed to provide evidence on current cycling levels and future cycling potential nationwide across Ireland.

A key consideration for TII is to ensure both urban and rural Ireland were targeted for cycling infrastructure, as shown in @fig-dublin.
To this end, TII extensively reviewed activity on Irish Roads and how the road network was being used by commuters, cyclists, heavy goods vehicles and other road users across the Irish transport network.
The aim was for the tool to provide strong, national, systematic but locally-specific evidence to monitor cycling friendliness and safety in Ireland, to ensure strategic alignment with national, regional and local policies.
The open source and publicly available nature of the tool ensures more inclusive and evidence-based conversations around cycle network planning between all stakeholders in the transport planning process.

TII also requires evidence to support monitoring of growth in cycling, upgrades of existing cycling infrastructure, business case development and appraisal, and exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive, for reporting collision rates of vulnerable road users by 2024.
These factors mean that the tool can be seen as an open access 'leverage point', providing key information for and supporting ambitious plans in many aspects the planning system, from network design to post-build monitoring [@lovelace2020].

::: {#fig-dublin}
![](images/paste-1.png)

![](images/paste-3.png)

Illustration of the density of route networks generated by the CRUSE Tool for Dublin City and surroundings (top) and rural County Mayo on the West coast of Ireland.
Source: CRUSE Tool, available at <https://cruse.bike/>.
:::

This paper aims to describe the design, features and potential use of the CRUSE Tool.
In @sec-methods, we outline the methods used to generate the evidence presented in the CRUSE Tool.
In @sec-results, we present the results of the CRUSE Tool, including estimates of current cycling levels and future cycling potential at the national, regional and local levels.
In @sec-discussion, we discuss the implications of the results for policy and practice.
In @sec-conclusions, we conclude the paper.

@@

The methods used to generate the evidence presented in the CRUSE Tool build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionized strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].

@@

Methods were developed to overcome each of these, as outlined in @sec-trip-purposes to @sec-ui.

<!-- Todo: add section on input datasets -->

@@

A feature of active travel interventions is that they require dense networks of routes to be effective [@parkin2018].
This means that high levels of geographic resolution are needed in the data used to estimate cycling potential.
However, datasets on travel patterns are often only available at the level of administrative zones.
The Central Statistics Office (CSO) in Ireland provides Place of Work, School or College Census of Anonymized Records ([POWSCAR](https://www.cso.ie/en/census/census2016reports/powscar/)) data on the number of people traveling to work and school at the Electoral Division (ED) level, for example.

In the PCT, the method used to convert OD data to route networks was to calculate a single route between the population weighted centroids of the zones associated with each OD pair.
This method works fine when the OD data represents movement between small areas, but was not appropriate for generating route networks from the POWSCAR data because zone centroids are so far apart that the resulting route networks would be sparse and unrealistic.
To tackle this issue we developed a new method for OD data disaggregation called 'jittering' [@lovelace2022].
The method works by first disaggregating the OD data based on a 'disaggregation threshold' and then assigning each disaggregated 'sub-OD' pair to 'subpoints' within each zone.
As shown in @fig-dublin, the resulting route networks are dense, even in rural areas.

@@

However, travel to school and work only constitute around only 30% of all trips in Ireland.
To overcome this issue, we developed a spatial interaction modeling methodology to estimate the number of trips between each OD pair for additional trip purposes.
The classification of trip purposes used in the CRUSE Tool was guided mainly by the trip purpose classification found within the National Household Travel Survey (NHTS), but with the addition of categories based on the comprehensive POWSCAR data, and the need to include recreational trips and multi-stage trips (not yet implemented).
An overview of the trip purposes used in CRUSE is presented in @fig-trip-purposes.

@@

CycleStreets' services can be used by anyone for free, but we commissioned a bespoke routing service for CRUSE, to enable:

@@

We calculated three network types for each disaggregated ('jittered') OD pair: 'Fastest', 'Balanced' and 'Quietest'.
As outlined on the CycleStreets' [website](https://www.cyclestreets.net/help/journey/howitworks/), the 'Fastest' routes minimize journey time, accounting for traffic lights and surface type.
The 'Quietest' route type minimizes busy sections of road, allowing high 'diversion factors' from the fastest route on quieter but less direct ways (often avoiding roads and interactions with motor vehicles altogether where possible).
The 'Balanced' route type is a compromize between the two, minimising journey time while avoiding the busiest roads.
Allowing users to switch between these three route types enables them to consider the trade-offs between directness and 'cycle friendliness' when planning new infrastructure, as shown in @fig-route-types.

@@

These statistics can be revealing: in Kildare it suggests that around half of all cycling activity occurs on network segments that are hostile (with a high inferred level of traffic stress), even when efforts are taken to avoid busy segments.
For the Fastest network, which may be more realistic for utility cycling, the proportion of cycling on hostile segments is even higher at 75%.

@@

Illustration of the Quietest (left) and Fastest (right) route networks for Naas (top) and Newbridge (bottom).
:::

@@

The Climate Action Plan scenario is loosely based on the Irish Government's [Climate Action Plan 2021](https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/), which contains policies for action to achieve a 51% reduction in overall GHG emissions by 2030, on a path towards net-zero by 2050.
For transport, this includes 500,000 extra walking, cycling and public transport trips per day by 2030.
In terms of car travel, the target is to "Increas\[e\] the proportion of kilometers driven by passenger electric cars to between 40 and 45% by 2030, in addition to a reduction of 10% in kilometers driven by the remaining internal combustion engine cars." This equates to a 5.5 to 6% reduction in total car km driven.

To model this decrease in car driver km, cycling uptake increases in line with the Go Dutch scenario.
However, we only model shift from driving to cycling.
There is no shift from other modes of transport to cycling.

@@

::: {#fig-user-stories}

![](images/mermaid-figure-1.png)

Hypothetical user stories for the CRUSE Tool illustrating that it can be used in different ways by different user groups.

:::

@@
